---
title: "Biostatistics - Task  1"
author: "Mauricio Marcos Fajgenbaun"
date: "13/5/2021"
output: html_document
---


```{r}
library( vcd )
library( dplyr )
library( mosaic )
```


## Exercise 1


```{r}
x <- data.frame("Density" = c("None","Low","Intermediate","High"), "Mean_Seagrass" = c(34.81,33.13,28.33,15.00))
x
x$Density <- (x$Density);
x$Density
```


```{r}
res <- aov(formula = Mean_Seagrass ~ Density, data = x)
anova(res)
```

```{r}
pairwise.t.test(x$Mean_Seagrass,x$Density , p.adj = " none ")
```
```{r}
TukeyHSD(res)
```

## Exercise 2




```{r}
A <- c(324,275,349,604,566,810,340,295,357,580,344,655,380,503,314)
B <- c(558,108,291,863,303,640,358,503,646,689,250,540,630,190)
```

First, let´s see if they behave normal or not.

```{r}
plot(density(A))
plot(density(B))
```

```{r}
shapiro.test(A)
shapiro.test(B)
```
First, we reject the null hypotesis of Group A having a normal distribution. Nevertheless, when we see the p-value of the second group (group B), it does not behave gaussian. When inspecting the density plots, we see that the group B does not seem to have a skewed distribution, as we can assume normality.

## Exercise 3

This is a categorical data problem. Now we want to do a comparison between to independent groups, the one that got conventional therapy and other that got  alternative therapy, and check the followeing hypotesis test:

Ho) The relapses and the groups are independent (there is homogeneity in the distribution of the relapses among the comparison groups).
H1) There is a difference in the distribution of the relapses among the two groups

The test is called Chi-squared test of independance:

```{r}
therapy = matrix ( c (2 ,8 ,21 ,16) ,
nrow = 2 ,
dimnames = list ( " Treatment " = c ( " Conventional " , " Alternative " ) ,
" Relapses " = c ( " Yes " , " No " )))
therapy
```




As the test statistic works only for big samples, I need to check that the expected frequencies are at least 5.

```{r}
chisq.test (therapy)$ expected
```

As we can see, there is one cell that the expected count is smaller than 5. Nevertheless, it is very close to 5. For this reason I will do the analysis both with a chi-square test and with a Fisher Exact test.

```{r}
chisq.test(therapy)
```


```{r}
chisq.test ( therapy , simulate.p.value = TRUE , B =1000)
```

With Fisher Exact test:

```{r}
fisher.test ( therapy)
```

